Report on laboratory assignments 1 and 2 for EDA282, \\ Parallel Computer Organization and Design

### 2014-10-15 lab/report/include/lab1.tex
corrections

\section{Lab 1}
\label{sec:lab1}
In this laboratory assignment, we evaluate overheads in cache coherence protocols using microbenchmarks. Three diverse benchmarks, described in section \ref{sec:lab11}, are used to compare the coherence protocols. In section \ref{sec:lab12}, MSI and MESI is compared and in section \ref{sec:lab13}, MESI and MESI-MG is compared. In each comparison, we use three working set: the first is smaller than an private L1 cache (512 counters), the second is larger than an private L1 cache and smaller than the LLC (1536 counters), and the last is larger than the LLC (98304 counters). The counters are stored in an array, the counter array, which is accessed by the benchmark.

\subsection{Task 1}
\label{sec:lab11}
Three different microbenchmarks with disgustingly various communication patterns are used in the simulations and here follow a description of each.
\subsubsection*{Benchmark 1 - Producer consumer}
In this benchmark, one thread increments the values in the counter array, whereas the other threads read these values and accumulate the read values in a private variable. However, it is not ensured that the incrementing thread increments the values before the other threads reads, and thus can this happen in any order. A counter access is done in mutual exclusion using a lock, and before a new lap of increments \& reads there is a barrier.

\subsubsection*{Benchmark 2 - No sharing}
This benchmark have no sharing of the counter array. The counter array is divided up equally among the threads, giving each thread access to every nth counter element, where n is the number of threads. Each thread then increments it's assigned counters sequentially. A barrier is placed before a new incrementation lap of the assigned array elements begin.

@@

In the no sharing benchmark the first read access to every element end in state S for MSI and state E for MESI, respectively. The following write transitions the element to state M for both the CCPs, but MSI have to issue a BusUpdate whereas MESI's transition is costless. The BusUpdate broadcasts an invalidate signal which gives MSI a slight reduction in performance compared to MESI. In the small working set the elements will stay in state M to the end of execution, but in working sets larger than a private L1 cache the element has to be fetched from higher levels. This makes MSI issue an invalidate broadcast for each increment and the accumulated overhead is easily spotted in \reffig{fig:lab1bars}.

@@

In the all shared benchmark, every thread increments every data element in mutual exclusion which favors MESI-MG. Each data element is read and modified exclusively by one processor at a time, thus is the S state not beneficial. As MESI-MG do not utilize the S state, as described, it is more suitable than MESI in this benchmark. In Table \ref{tbl:lab1:task3} it is shown that the CCPs results are alike, except that MESI have to issue a significant number of invalidates, which is needed for the transition from state S to M. As shown in \reffig{fig:lab1bars}, MESI-MG's benefit is not visible for the working set size increasing the size of the LLC. That is because the cycle count is dominated by the LLC misses, as in the producer consumer benchmark.

>>>>

\section{Laboratory assignment 1}
\label{sec:lab1}
In this laboratory assignment, we evaluate overheads in cache coherence protocols using microbenchmarks. Three diverse benchmarks, described in section \ref{sec:lab11}, are used to compare the coherence protocols. In section \ref{sec:lab12}, MSI and MESI are compared and in section \ref{sec:lab13}, MESI and MESI-MG are compared. In each comparison, we use three working set: the first is smaller than an private L1 cache (512 counters), the second is larger than an private L1 cache and smaller than the LLC (1536 counters), and the last is larger than the LLC (98304 counters). The counters are stored in an array, the counter array, which is accessed by the benchmark.

\subsection{Task 1}
\label{sec:lab11}
Three different microbenchmarks with disgustingly various communication patterns are used in the simulations, and here follow a description of each.
\subsubsection*{Benchmark 1 - Producer consumer}
In this benchmark, one thread increments the values in the counter array, whereas the other threads read these values and accumulate the read values in a private variable. However, it is not ensured that the incrementing thread increments the values before the other threads reads, and thus can this happen in any order. A counter access is done in mutual exclusion using a lock, and before a new lap of increments \& reads there is a barrier.

\subsubsection*{Benchmark 2 - No sharing}
This benchmark has no sharing of the counter array. The counter array is divided up equally among the threads, giving each thread access to every nth counter element, where n is the number of threads. Each thread then increments it's assigned counters sequentially. A barrier is placed before a new incrementation lap of the assigned array elements begin.

@@

In the no sharing benchmark the first read access to every element end in state S for MSI and state E for MESI, respectively. The following write transitions the element to state M for both the CCPs, but MSI has to issue a BusUpdate whereas MESI's transition is costless. The BusUpdate broadcasts an invalidate signal which gives MSI a slight reduction in performance compared to MESI. In the small working set the elements will stay in state M to the end of execution, but in working sets larger than a private L1 cache the element has to be fetched from higher levels. This makes MSI issue an invalidate broadcast for each increment and the accumulated overhead is easily spotted in \reffig{fig:lab1bars}.

@@

In the all shared benchmark, every thread increments every data element in mutual exclusion which favors MESI-MG. Each data element is read and modified exclusively by one processor at a time, thus is the S state not beneficial. As MESI-MG do not utilize the S state, as described, it is more suitable than MESI in this benchmark. In Table \ref{tbl:lab1:task3} it is shown that the CCPs results are alike, except that MESI has to issue a significant number of invalidates, which is needed for the transition from state S to M. As shown in \reffig{fig:lab1bars}, MESI-MG's benefit is not visible for the working set size increasing the size of the LLC. That is because the cycle count is dominated by the LLC misses, as in the producer consumer benchmark.
